---
title: "R Coding Exercise"
---

### **Module 3 - R Coding Exercise**

**Done by: Shaun van den Hurk**

Getting started and set up with packages and retrieving dataset.

```{r}
#Installing required packages ("dslabs", "tidyverse")
options(repos = c(CRAN = "https://cran.rstudio.com/")) #Attempting to address issue with downloading dslabs script at render
install.packages("dslabs") #for dataset
install.packages("tidyverse") #for tools for analysis

#load required libraries ("dslabs", "tidyverse", "ggplot2", "here")
library("dslabs")
library("tidyverse")
library("ggplot2") #for generated plots

```

#Getting an overview of the data

#Look at the gapminder data help file help(gapminder)

```{r}
#Get an overview of the gapminder data structure
str(gapminder)

#Get a summary of the gapminder data
summary(gapminder)

```

#Determine the type of the object gapminder

class(gapminder)

######################### 

#Getting started with data processing

#Filter out data from African countries by filtering by continent

```{r}
#Filter out African data
africadata <- gapminder |> filter(continent =="Africa")
#View the newly filtered data
View(africadata)
```

#From the filtered Africa data, create two new objects, one with the two columns "infant_mortality" and "life_expectancy", and second object with the two columns "population" and "life_expectancy".

#Creating the first object named "africa_1" and looking at the overview and summary of the newly created object

```{r}
#Using the "select" function to select the columns from the previous "africadata" dataset
africa_1 <- africadata |> select(infant_mortality, life_expectancy)
#get an overview of the new africa_1 datset
str(africa_1)
#see a summary of the new africa_1 dataset
summary(africa_1)
#View the resulting object more closely
view(africa_1)

```

#Creating the second object named "africa_2" with the columns "population" and "life_expectancy" following the same approach as before

```{r}
#Using the "select" function to select the columns from the previous "africadata" dataset
africa_2 <- africadata |> select(population, life_expectancy)
#get an overview of the new africa_1 datset
str(africa_2)
#see a summary of the new africa_1 dataset
summary(africa_2)
#View the resulting object more closely
view(africa_2)

```

#Create two different plots to evaluate the data from the objects arica_1 and africa_2 using the ggplot2 package - previously loaded.

#Generate a plot demonstrating life expectancy as a function of infant mortality.

```{r}
#Generate the plot as a scatter plot with the variables associated with the approiate axes, reduce point size to 1, set colour to blue, label different areas of the graph
ggplot(africa_1, aes(x = infant_mortality, y = life_expectancy)) +   
  geom_point(size = 1, color = "blue") +
  labs(title = "Infant mortality vs Life expectancy", x = "Infant mortality", y = "Life expectancy" ) 

#Save the generated plot
ggsave("africa_1_Life_expectancy_vs_infant_mortality.png", width = 8, height =6)
```

The data shows shows a negative correlation between infant mortality and life expectancy. This is intuitive.

#Generate a second plot demonstrating life expectancy as a function of population size, where the x-axis (population) is plotted on a log scale

```{r}
#Generate the plot as a scatter plot with the variables associated with the approiate axes,set the x-axis (population) to be in the log-scale, reduce point size to 1, set colour to red, label different areas of the graph
ggplot(africa_2, aes(x = population, y = life_expectancy)) +   
  geom_point(size = 1, color = "red") +
  scale_x_log10() +
  labs(title = "Population vs Life expectancy", x = "Population (Log-scale)", y = "Life expectancy" ) 

#Save the generated plot
ggsave("africa_2_Life_expectancy_vs_population_log_scale.png", width = 8, height =6)
```

The plots generated are a bit messy and difficult to interpret as they currently are. The strange appearance is most likely a result of the repition of the data (collected data) that has occured by reporting the variables from 1962 to 2016. Therefore, there is a great deal of overlap and it is hard to see any real relationships or trends. This is amplified by the fact that the life expectancy and certainly the populaion changes each year and so the points are being shifted. It might be best to focus on a particular country or region over a period of time, or to focus on one particular year to reduce this overlap. T

#Search for years with missing (NA) data points for infant mortality.

```{r}
#I used the textbook and ChatGPT to help me generate and correct my code. I tried the select function with & first which is not apprpriate for rows
missing_infant_mortality <- africadata |> #Assigning new variable and searching in the dataset
filter(is.na(infant_mortality)) |>  #continuing with pipe and filtering for where infant mortality is "NA"
select(year)  #continuing pipe and keeping the years where infant mortality is NA"

```

#Create a new object "africa_3_y2000" with the extracted data from the year 2000 from "africadata" and view the summary of the data

```{r}
#Use the filter function to select the columns where the year is 2000
africa_3_y2000 <- africadata |> filter(year==2000)
#view the data structure of the new object africa_3_y2000
str(africa_3_y2000)
#view a summary of the new object africa_3_y2000
summary(africa_3_y2000)
```

#Make new plots for the data from the year 2000 following the same approach and similar code as before

#Generate a plot to view infant mortality vs life expectancy from the year 2000 by using the filtered object created for the year 2000 from the Africa dataset that was originally created.

```{r}
#Generate the plot as a scatter plot with the variables associated with the approiate axes, reduce point size to 1, set colour to black, label different areas of the graph
ggplot(africa_3_y2000, aes(x = infant_mortality, y = life_expectancy)) +   
  geom_point(size = 1, color = "black") +
  labs(title = "Infant mortality vs Life expectancy from the year 2000", x = "Infant mortality", y = "Life expectancy" ) 

#Save the generated plot
ggsave("africa_3_y2000_Life_expectancy_vs_infant_mortality.png", width = 8, height =6)
```

#Generate a simialr plot to view population (on a log-scale) vs life expectancy from the year 2000 by using the filtered object created for the year 2000 from the Africa dataset that was originally created.

```{r}
#Generate the plot as a scatter plot with the variables associated with the approiate axes,set the x-axis (population) to be in the log-scale, reduce point size to 1, set colour to red, label different areas of the graph
ggplot(africa_3_y2000, aes(x = population, y = life_expectancy)) +   
  geom_point(size = 1, color = "red") +
  scale_x_log10() +
  labs(title = "Population vs Life expectancy for the year 2000", x = "Population (Log-scale)", y = "Life expectancy" ) 

#Save the generated plot
ggsave("africa_3_y2000_Life_expectancy_vs_population_log_scale.png", width = 8, height =6)
```

#Fit a linear model to the data to help with further interpretation. Use the "lm" function to fit life expectancy as the outcome, with infant mortality as the predictor. This is based on the data from the year 2000 in Africa only.

```{r}
fit1 <- lm(life_expectancy ~ infant_mortality, data = africa_3_y2000)
#View the summary of the model
summary(fit1)

```

From this fit it appears that infant moratlity has a strong and highly significant association with life expectancy (p value is 2.83e-08). We see that higher infant mortality is associated with a lower life expectancy.

#Generate a similar linear fit with life expectancy as the outcome and population size as the predictor. This is based on the data from the year 2000 in Africa only.

```{r}
fit2 <- lm(life_expectancy ~ population, data = africa_3_y2000)
#View the summary of the model
summary(fit2)

```

There does not appear to be any statistical significance in the relationship between population size and life expectancy in this dataset modelled. The p-value was 0.6159, which is not significant. This indicates that population size is not an effective predictor of life expectancy in this dataset.





```{r}
<!-- This section contributed by Asmith Joseph -->

```



```{r}
# Taking a look at the dslabs to identify which dataset I want to choose for the assignment
library(dslabs)
data(package = "dslabs")

```

```{r}
# I chose us_contagious_diseases dataset
# Loading the  us_contagious_diseases datasetthe Dataset
library(dslabs)
data("us_contagious_diseases")

```


```{r}
#The dataset contains the following columns:

#disease: The name of the disease (e.g., "Measles," "Polio").
#state: The U.S. state where the data was recorded.
#year: The year the data was reported.
#weeks_reporting: The number of weeks during the year in which the state reported data.
#count: The number of reported cases of the disease.
#population: The population of the state in the respective year.
#rate: The number of disease cases per 10,000 people
```


```{r}
#Exploring the datasets to identify variables and so on 
head(us_contagious_diseases)


#checking the structure 
str(us_contagious_diseases)


# Exploring the summary 
summary(us_contagious_diseases)

```



```{r}
#Processing and cleaning the Data. First I am filtering out diseases with missing data, mostly focusing on one specific diseases, Measles. 
### Data Processing I focus on measles data for this analysis. The dataset is cleaned to remove rows with missing values, and a new variable, `rate_per_100k`, is calculated to represent cases per 100,000 people.


# filtering out the data for measles
measles <- us_contagious_diseases %>%
  filter(disease == "Measles") %>%
  drop_na()  # Remove rows with missing values

# Adding a column for the cases per 100,000 population
measles <- measles %>%
  mutate(rate_per_100k = (count / population) * 100000)

# Previewing the cleaned dataset
head(measles)

# Checking for missing values
sum(is.na(measles))

```





```{r}
### In the next part, I am doing exploratory Figures by visualizing the trends of measles cases over time and across states. 1) figure shows the number of cases over the years, 2) second figure is a heatmap of cases by state and year.
```


```{r}
#Creating exploratory figures, such as Visualize trends, distributions, or summaries.

# Plot the total number of measles cases over time
ggplot(measles, aes(x = year, y = count)) +
  geom_line(color = "pink") +
  labs(title = "Measles Cases Over Time",
       x = "Year",
       y = "Number of Cases") +
  theme_minimal()

```

```{r}
#  2nd figure Measles Cases by State (Heatmap)
# Creating a heatmap of cases by state and year
measles_heatmap <- measles %>%
  group_by(state, year) %>%
  summarize(total_cases = sum(count, na.rm = TRUE))

ggplot(measles_heatmap, aes(x = year, y = reorder(state, total_cases), fill = total_cases)) +
  geom_tile(color = "white") +
  scale_fill_gradient(low = "white", high = "red") +
  labs(title = "Measles Cases Heatmap by State and Year",
       x = "Year",
       y = "State",
       fill = "Total Cases") +
  theme_minimal()

```




```{r}
Next Part, I am focusing on Statistical Model. fitting a simple linear regression model to examine the trend of measles cases over time. The model uses `year` as the predictor and `count` (number of cases) as the outcome variable.
```


```{r}
# Fitting a linear model to examine the trend of measles cases over time
measles_lm <- lm(count ~ year, data = measles)

# Summarizing the model
summary(measles_lm)

```

```{r}
### Results from the Linear Model
#The summary of the linear regression model shows the following:
# 1) The slope coefficient for `year` is negative, indicating a decline in measles cases over time, 2) The model's p-value suggests that this decline is statistically significant.

```



